---
title: "Assignment 1"
author: "Yuxin Sun"
date: "07/15/2025"
header-includes:
spacing: double
output:
  pdf_document: default
  html_document:
    mathjax: default
  word_document: default
always_allow_html: true
citation_package: natbib
#bibliography: ["research sources.bib"]     
biblio-style: "apalike"
link-citations: true
nocite: | 
  @SIRcode, @FB, @OB, @Article,@Code, @MKMK
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Q1(a)

$$
\text{Let the linear model be: } \quad \mathbf{y} = M_Y \boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

- $ \mathbf{y} = [y_1, y_2, \ldots, y_8]^T $: vector of scale readings from the 8 weighings  
- $ \boldsymbol{\beta} = [\beta_1, \beta_2, \ldots, \beta_7]^T $: true weights of the 7 light objects  
- $ M_Y $: design matrix under Yates’ procedure indicating object presence in each weighing  
- $ \boldsymbol{\varepsilon} = [\varepsilon_1, \varepsilon_2, \ldots, \varepsilon_8]^T $: independent random errors, each with mean zero and variance $ \sigma^2 $




$$
\mathbf{y} = M_Y \boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

$$
\begin{bmatrix}
y_1 \\
y_2 \\
y_3 \\
y_4 \\
y_5 \\
y_6 \\
y_7 \\
y_8 \\
\end{bmatrix}
=
\begin{bmatrix}
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 0 & 0 & 0 & 0 \\
1 & 0 & 0 & 1 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 & 0 & 1 & 1 \\
0 & 1 & 0 & 1 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 & 1 & 0 & 1 \\
0 & 0 & 1 & 1 & 0 & 0 & 1 \\
0 & 0 & 1 & 0 & 1 & 1 & 0 \\
\end{bmatrix}
\begin{bmatrix}
\beta_1 \\
\beta_2 \\
\beta_3 \\
\beta_4 \\
\beta_5 \\
\beta_6 \\
\beta_7 \\
\end{bmatrix}
+
\begin{bmatrix}
\varepsilon_1 \\
\varepsilon_2 \\
\varepsilon_3 \\
\varepsilon_4 \\
\varepsilon_5 \\
\varepsilon_6 \\
\varepsilon_7 \\
\varepsilon_8 \\
\end{bmatrix}
$$

---


The matrix $ M_Y \in \mathbb{R}^{8 \times 7} $ represents which items are used in each of the eight weighings. Each row corresponds to one weighing, and each column to an object.

---
```{r}
matrix_M_Y <- matrix(c(
  1,1,1,1,1,1,1,
  1,1,1,0,0,0,0,
  1,0,0,1,1,0,0,
  1,0,0,0,0,1,1,
  0,1,0,1,0,1,0,
  0,1,0,0,1,0,1,
  0,0,1,1,0,0,1,
  0,0,1,0,1,1,0
), nrow = 8, byrow = TRUE)

# Compute the projection matrix for least squares estimation
proj_Y <- solve(t(matrix_M_Y) %*% matrix_M_Y) %*% t(matrix_M_Y)
proj_Y
```

---

#### Least Squares Estimator

$$
\hat{\boldsymbol{\beta}} = \left( M_Y^{\top} M_Y \right)^{-1} M_Y^{\top} \mathbf{y}
$$

This is the ordinary least squares estimator for the object weights under Yates’ method, assuming homoscedastic and independent errors.



### Q1(b)

$$
\text{Now consider the modified procedure proposed by Hotelling, where: } \quad \mathbf{y} = M_H \boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

- $ \mathbf{y} = [y_1, y_2, \ldots, y_8]^T $: net scale readings from Hotelling's configuration  
- $ \boldsymbol{\beta} = [\beta_1, \beta_2, \ldots, \beta_7]^T $: true weights of the seven objects  
- $ M_H $: Hotelling’s design matrix, where the remaining objects are placed in the opposite pan of the balance scale, leading to $\pm 1$ contrast coding  
- $ \boldsymbol{\varepsilon} = [\varepsilon_1, \varepsilon_2, \ldots, \varepsilon_8]^T $: independent noise terms with zero mean and constant variance

---

#### Matrix Formulation

$$
\mathbf{y} = M_H \boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

$$
\begin{bmatrix}
y_1 \\
y_2 \\
y_3 \\
y_4 \\
y_5 \\
y_6 \\
y_7 \\
y_8 \\
\end{bmatrix}
=
\begin{bmatrix}
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & -1 & -1 & -1 & -1 \\
1 & -1 & -1 & 1 & 1 & -1 & -1 \\
1 & -1 & -1 & -1 & -1 & 1 & 1 \\
-1 & 1 & -1 & 1 & -1 & 1 & -1 \\
-1 & 1 & -1 & -1 & 1 & -1 & 1 \\
-1 & -1 & 1 & 1 & -1 & -1 & 1 \\
-1 & -1 & 1 & -1 & 1 & 1 & -1 \\
\end{bmatrix}
\begin{bmatrix}
\beta_1 \\
\beta_2 \\
\beta_3 \\
\beta_4 \\
\beta_5 \\
\beta_6 \\
\beta_7 \\
\end{bmatrix}
+
\begin{bmatrix}
\varepsilon_1 \\
\varepsilon_2 \\
\varepsilon_3 \\
\varepsilon_4 \\
\varepsilon_5 \\
\varepsilon_6 \\
\varepsilon_7 \\
\varepsilon_8 \\
\end{bmatrix}
$$

---

#### Design Matrix in Hotelling’s Method ($M_H$)

This matrix $ M_H \in \mathbb{R}^{8 \times 7} $ encodes weighings such that the group of objects not included in a weighing are placed in the **opposite pan**. This results in signed contrasts of weights across the scale.

---

#### R Code

```{r}
# Define design matrix for Hotelling's weighing method
matrix_M_H <- matrix(c(
   1, 1, 1, 1, 1, 1, 1,
   1, 1, 1,-1,-1,-1,-1,
   1,-1,-1, 1, 1,-1,-1,
   1,-1,-1,-1,-1, 1, 1,
  -1, 1,-1, 1,-1, 1,-1,
  -1, 1,-1,-1, 1,-1, 1,
  -1,-1, 1, 1,-1,-1, 1,
  -1,-1, 1,-1, 1, 1,-1
), nrow = 8, byrow = TRUE)

# Compute least squares estimator matrix
proj_H <- solve(t(matrix_M_H) %*% matrix_M_H) %*% t(matrix_M_H)
proj_H
```

---

#### Least Squares Estimator

$$
\hat{\boldsymbol{\beta}} = \left( M_H^{\top} M_H \right)^{-1} M_H^{\top} \mathbf{y}
$$

---

#### Componentwise Form of Estimator

$$
\begin{aligned}
\hat{\beta}_1 &= 0.125(y_1 + y_2 + y_3 + y_4 - y_5 - y_6 - y_7 - y_8) \\
\hat{\beta}_2 &= 0.125(y_1 + y_2 - y_3 - y_4 + y_5 + y_6 - y_7 - y_8) \\
\hat{\beta}_3 &= 0.125(y_1 + y_2 - y_3 - y_4 - y_5 - y_6 + y_7 + y_8) \\
\hat{\beta}_4 &= 0.125(y_1 - y_2 + y_3 - y_4 + y_5 - y_6 + y_7 - y_8) \\
\hat{\beta}_5 &= 0.125(y_1 - y_2 + y_3 - y_4 - y_5 + y_6 - y_7 + y_8) \\
\hat{\beta}_6 &= 0.125(y_1 - y_2 - y_3 + y_4 + y_5 - y_6 - y_7 + y_8) \\
\hat{\beta}_7 &= 0.125(y_1 - y_2 - y_3 + y_4 - y_5 + y_6 + y_7 - y_8) \\
\end{aligned}
$$

### Q1(c)

We now compare the variance of each estimated weight under the two methods by using their coefficient structures.

---


From part (a), the estimator for the first object weight looks like:

$$
\hat{\beta}_1 = 0.0625\,y_1 + 0.3125(y_2 + y_3 + y_4) - 0.1875(y_5 + y_6 + y_7 + y_8)
$$

Since all observations are independent with constant variance $\sigma^2$, the variance of $\hat{\beta}_1$ becomes:

$$
\text{Var}(\hat{\beta}_1) = (0.0625)^2 + 3 \cdot (0.3125)^2 + 4 \cdot (0.1875)^2
$$

Computing the components:

$$
0.00390625 + 3 \cdot 0.09765625 + 4 \cdot 0.03515625 = 0.00390625 + 0.29296875 + 0.140625 = 0.4375
$$

So under Yates’ configuration:

$$
\text{Var}(\hat{\beta}_i) = 0.4375\,\sigma^2 \quad \text{for all } i
$$

---


From part (b), the estimate for $\beta_1$ is:

$$
\hat{\beta}_1 = 0.125(y_1 + y_2 + y_3 + y_4 - y_5 - y_6 - y_7 - y_8)
$$

Each coefficient is either $+0.125$ or $-0.125$, and there are 8 of them. So the variance is:

$$
\text{Var}(\hat{\beta}_1) = 8 \cdot (0.125)^2 = 8 \cdot 0.015625 = 0.125
$$

Therefore, under Hotelling’s design:

$$
\text{Var}(\hat{\beta}_i) = 0.125\,\sigma^2 \quad \text{for all } i
$$
### Q1(d)
- Yates:$\text{Var}(\hat{\beta}_i) = 0.4375\,\sigma^2$
- Hotelling:$\text{Var}(\hat{\beta}_i) = 0.125\,\sigma^2$

Since $0.125 < 0.4375$, the Hotelling procedure yields smaller variances for the estimated weights, meaning more accurate estimates of the object weights.
Therefore, if the chemist's goal is to achieve the highest possible precision, then Hotelling’s method is the better choice.

\newpage

### Q2(a)

We are given the yields from six bean plants, with three plants assigned to each fertilizer group:

- Fertilizer A: $[3, 3, 6]$
- Fertilizer B: $[5, 4, 7]$

To construct the randomization distribution, we assume the labels A and B were randomly assigned. We combine all six values and compute the difference in group means for all possible allocations of 3 to A and 3 to B.


```{r}
# Combine all yields 
yield_all <- c(3, 3, 6, 5, 4, 7)

# Get all unique ways to select 3 observations for one group
perm_matrix <- combn(1:6, 3)

# For each combination, calculate difference in group means
rand_dist <- apply(perm_matrix, 2, function(group_A) {
  group_B <- setdiff(1:6, group_A)
  mean(yield_all[group_A]) - mean(yield_all[group_B])
})

# Display the randomization distribution
rand_dist
```

This vector `rand_dist` contains all possible values of the difference in means under the null hypothesis of no treatment effect.



### Q2(b)

We calculate the p-value using the randomization distribution. We compare the observed difference in means to the null distribution created in part (a).


```{r}
# Observed difference in means
obs_diff <- mean(c(3, 3, 6)) - mean(c(5, 4, 7))

# Null center (mean of permutation distribution)
null_mean <- mean(rand_dist)

# Two-sided p-value using permutation definition
p_val <- mean(abs(rand_dist - null_mean) >= abs(obs_diff - null_mean))

round(p_val, 2)
```

This p-value represents the proportion of randomized allocations that produce a difference in means as or more extreme than the observed one.

---

### Q2(c)

We compare the above to a two-sample t-test, assuming equal variance.


```{r}
group_A <- c(3, 3, 6)
group_B <- c(5, 4, 7)

t.test(group_A, group_B, var.equal = TRUE)
```

This test returns a t-statistic, degrees of freedom, and a p-value under the assumption of normality and equal variances in the two groups.

---

### Q2(d)

The p-value from the randomization test was approximately 0.50, and the p-value from the t-test was approximately 0.37.

Both results suggest that the difference in yields between Fertilizer A and Fertilizer B is not statistically significant.

Thus, both approaches lead to the same practical conclusion:  
There is no clear evidence that the two fertilizer types result in different yields for bean plants under this experimental setup.

\newpage

### Q3(a)

The sequence of coin flips given is:

```
T T H T H T T T H T T H
```

We apply the rule:  
- If the result is H, assign Drug A to the right eye  
- If the result is T, assign Drug A to the left eye

Below is the resulting treatment assignment for each subject:

$$
\begin{array}{|c|c|c|c|}
\hline
\textbf{Subject} & \textbf{Left Eye} & \textbf{Right Eye} & \textbf{Coin Flip} \\
\hline
1 & A & B & T \\
2 & A & B & T \\
3 & B & A & H \\
4 & A & B & T \\
5 & B & A & H \\
6 & A & B & T \\
7 & A & B & T \\
8 & A & B & T \\
9 & B & A & H \\
10 & A & B & T \\
11 & A & B & T \\
12 & B & A & H \\
\hline
\end{array}
$$

---

### Q3(b)

Each flip of the coin is an independent Bernoulli trial with a probability of 0.5. Since there are 12 subjects and hence 12 independent coin flips, the probability of getting this specific sequence of heads and tails is:

$$
P(\text{exact sequence}) = \left( \frac{1}{2} \right)^{12} = \frac{1}{4096}
$$

This reflects the uniform probability of any single assignment sequence in this randomization scheme.

---

### Q3(c)

This experiment uses a randomized paired design.

Each subject contributes two experimental units: one left eye and one right eye. The treatment is randomized within each subject using a fair coin flip to decide which eye receives Drug A and which receives Drug B.

This type of design has the following benefits:

- Controls for subject-level variability 
- Ensures that each treatment is compared within-subject, not between different people
- Leads to more precise estimation of treatment effects with fewer experimental units

Such within-unit randomization is ideal for paired anatomical structures like eyes, ears, or limbs.

\newpage

### Q4(a)

We want to determine whether the promotional strategy had an effect on hotwing consumption. There are 30 participants, randomly split into two groups:

- Group M: received marketing pitch + beer coupon  
- Group F: standard customers (no promotion)

Each group contains 15 subjects, and we compare their mean hotwing intake.

The number of possible group assignments is:

$$
\binom{30}{15} = 155,\!117,\!520
$$

This makes exhaustive enumeration infeasible. Instead, we approximate the null distribution of the mean difference using Monte Carlo simulation.



```{r}
library(resampledata)
data(Beerwings)
set.seed(123)

hotwings <- Beerwings$Hotwings
condition <- Beerwings$Gender

# Observed difference: mean M - mean F
obs_diff <- mean(hotwings[condition == "M"]) - mean(hotwings[condition == "F"])

# Approximate null distribution
n_iter <- 250000
simulated_diffs <- numeric(n_iter)

for (i in 1:n_iter) {
  sampled <- sample(30, 15)
  simulated_diffs[i] <- mean(hotwings[sampled]) - mean(hotwings[-sampled])
}

# Monte Carlo p-value (two-sided)
p_val <- (sum(abs(simulated_diffs) >= abs(obs_diff)) + 1) / (n_iter + 1)
round(p_val, 4)
```

---

### Q4(b)

This is an experiment, not an observational study.

Participants were randomly assigned to the treatment (promotion) or control (no promotion) groups. The researcher directly controlled group membership, which satisfies the definition of an experimental design.

---

### Q4(c)

We now perform a two-sample t-test to test the same hypothesis using traditional methods.


```{r}
t.test(Hotwings ~ Gender, data = Beerwings, var.equal = TRUE)
```

---

####️ Assumptions:

Both tests rely on:

- Independent observations  
- Only systematic difference is the treatment (due to randomization)

T-test also assumes:

- Normality (or large sample size)  
- Equal population variances in both groups

---

#### Result

The p-value from Monte Carlo simulation was approximately 0.0021, and the t-test p-value was about 0.0015.

Both methods indicate a statistically significant effect. The promotion increased hotwing consumption.

---

### Q4(d)

We now test whether the promotion also affected beer consumption.
#### Normality:

```{r}
beer_data <- Beerwings$Beer

qqnorm(beer_data[condition == "M"], main = "QQ Plot - Group M")
qqline(beer_data[condition == "M"])

qqnorm(beer_data[condition == "F"], main = "QQ Plot - Group F")
qqline(beer_data[condition == "F"])
```

Visual inspection reveals departures from normality, suggesting a randomization test may be preferable over a t-test.

#### Randomization Test (Beer)

```{r}
set.seed(123)

obs_diff_beer <- mean(beer_data[condition == "M"]) - mean(beer_data[condition == "F"])

sim_beer <- numeric(n_iter)

for (i in 1:n_iter) {
  group_A <- sample(30, 15)
  sim_beer[i] <- mean(beer_data[group_A]) - mean(beer_data[-group_A])
}

pval_beer <- (sum(abs(sim_beer) >= abs(obs_diff_beer)) + 1) / (n_iter + 1)
round(pval_beer, 4)
```

---

#### Result

The Monte Carlo p-value for beer consumption is approximately 0.0628.

This result is not statistically significant at the 5% level. We didn't find convincing evidence that the promotion impacted beer consumption.

---

### Q4(e)

If the researcher instead recruited 40 participants, assigning 30 to group M and 10 to group F, the total number of possible treatment allocations is:

$$
\binom{40}{30} = \binom{40}{10} = 847,\!660,\!528
$$

This number grows rapidly with sample size imbalance, which is why simulation-based inference is practical in randomized experiments.

